---
title: "data-management-2025"
author: "Moegamad Uzair Jack"
date: "2025-02-21"
output: html_document
---


## Loading packages 
Loading all necessary packages for tidy data and analyses

```{r}

library(readr)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(conflicted)
library(lubridate)
library(dplyr)
```

## Reading data
Loading data from csv file and viewing it
```{r}

RawSightingsData <- read_csv("RawSightingsData.csv")
View(RawSightingsData)

```
## Inspecting the structure of the data
Viewing the structure and summary of the data

```{r}

glimpse(RawSightingsData)
head(RawSightingsData)
summary(RawSightingsData)

```

## Column and Cell Cleaning
Manually clean column names using gsub and replace spaces with underscores and 
remove special characters
  
```{r}
colnames(RawSightingsData) <- gsub(" ", "_", colnames(RawSightingsData))  
colnames(RawSightingsData) <- gsub("[^A-Za-z0-9_]", "", 
                                   colnames(RawSightingsData))  
```
## Data Conversion
Convert 'date' column to Date format

```{r}
RawSightingsData <- RawSightingsData %>%
  mutate(Date = as.Date(Date, format = "%d/%m/%Y"))
```

## Print cleaned column names

```{r}
print(colnames(RawSightingsData))
```

## More Column and Cell Cleaning
Cleaning columns and assigning all invalid or missing data the value NA 

```{r}

columns_to_clean <- c("Size_Class", "Time", "No_Remora", "New_Injury", 
                      "Preg__FMW__YoY" )
RawSightingsData[columns_to_clean] <- RawSightingsData[columns_to_clean] %>%
  lapply(function(x) ifelse(x == "-", NA, x))

columns_to_clean <- c("Sex" )
RawSightingsData[columns_to_clean] <- RawSightingsData[columns_to_clean] %>%
  lapply(function(x) ifelse(x == "N/A", NA, x))
```

## Remove columns where all values are NA

```{r}
RawSightingsData <- RawSightingsData %>%
  select(where(~ !all(is.na(.))))

glimpse(RawSightingsData)
```
## Deleting redundant columns

```{r}
 RawSightingsData <- RawSightingsData %>% select(-`Source`, -`Source_Type`)

```
## LOng Format Conversion
Convert data to long format
```{r}
long_data <- RawSightingsData %>%
  pivot_longer(cols = c('Behavioural_Activity', 'No_Remora'),
               names_to = 'Observation_Type',
               values_to = 'Observation_Value')

long_data <- long_data %>%
  select(ID_No, Sex, Resight_YN, Date, Region, Observation_Value, Observation_Type, Source_Name)


view(long_data)
```
## Figure Plotting
Plotting a stacked bar graph of the Re-sight frequency by region
```{r}
resight_frequency <- long_data %>%
  group_by(Region, `Resight_YN`) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  mutate(Percentage = Count / sum(Count) * 100)

ggplot(resight_frequency, aes(x = Region, y = Count, fill = `Resight_YN`)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() +
  labs(title = "Re-sight Frequency by Region (Percentage)",
       x = "Region",
       y = "Percentage")
```

Plotting bar graph of the sex distribution by region
```{r}
sex_distribution <- long_data %>%
  group_by(Region, Sex) %>%
  summarise(Count = n(), .groups = 'drop')

ggplot(sex_distribution, aes(x = Region, y = Count, fill = Sex)) +
  geom_col(position = "dodge", width = 0.7) +
  facet_wrap(~Sex) +
  theme_minimal() +
  labs(title = "Sex Distribution by Region",
       x = "Region",
       y = "Count")

```




